---
title: "Data Access and Publishing"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data Access and Publishing}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

GitHub Releases are the data transport layer of the whole pannaverse: raw
Opta data, trained models, prediction caches, and every published rating all
move between repos as release assets, not as files in git. This vignette
covers pulling data down and, if you're changing a pipeline, publishing it
back out.

## Downloading Opta data: `pb_download_opta()`

`pb_download_opta()` is the canonical Opta puller -- an **incremental sync**,
not a one-shot tarball download. It compares local files against the
`opta-latest` release asset list and only pulls what's missing, size-changed,
or (optionally) remote-newer:

```{r download-opta}
library(panna)

# See what's out of date without downloading anything
pb_download_opta(dry_run = TRUE)

# Pull only the changed consolidated files (safe to re-run daily)
pb_download_opta()

# Restrict to a subset of assets by name pattern
pb_download_opta(pattern = "^events_")

# Force a full re-download regardless of local state
pb_download_opta(force = TRUE)
```

It works from an empty directory too, so it's also the fresh-clone puller --
`data-raw/bootstrap.R` calls it directly. (An older tarball-based downloader
was removed in 2026-07: `opta-latest` no longer carries an archive asset, only
individual consolidated parquets, so incremental per-asset sync is the only
path that works.)

```{r bootstrap-fresh-clone}
# Rscript data-raw/bootstrap.R            # Opta data + models + caches
# Rscript data-raw/bootstrap.R opta       # Opta data + models only
```

Downloading match predictions and pre-trained models works the same way,
with their own dedicated releases:

```{r download-other}
pb_download_predictions()   # predictions.parquet from predictions-latest
pb_download_epv_models()    # xg_model.rds / xpass_model.rds / epv_model.rds
```

`pb_download_epv_models()` pulls only those three files (not the WP or xDuel
models) from `repo = "peteowen1/pannadata"`, `tag = "epv-models"` by default
-- pass explicit `repo`/`tag` if you need a different model release. See
`../MODELS.md` for the full model inventory, which loader resolves which
file, and the silent-stale-fallback trap when a model source is missing.

## The release-as-data-bus pattern

Most releases live on `peteowen1/pannadata`; pre-trained models live on the
separate `peteowen1/pannamodels` repo instead. The ones you'll touch most:

| Release tag | Contents | Written by | Read by |
|-------------|----------|-----------|---------|
| `opta-latest` | Consolidated Opta parquets + per-league events + xMetrics | Daily Opta scrape; `epv-pipeline.yml` for xMetrics | `pb_download_opta()`, all `load_opta_*()` loaders |
| `epv` (on `pannamodels`) | `xg_model.rds`, `epv_model.rds`, `wp_model.rds`, `duel_model.rds`, etc. | Manual, one-time per retrain | `pb_download_epv_models()`, `load_*_model()` family |
| `predictions-cache` | RAPM + Skills `.rds` caches | `upload_prediction_caches.R` | Predictions pipeline (GHA) |
| `predictions-latest` | `predictions.parquet`/`.csv` | Predictions pipeline step 13 | `pb_download_predictions()`, `load_predictions()` |
| `blog-latest` | `panna_ratings.parquet`, `game_logs_*.parquet`, `action_equity_*.parquet` | Predictions pipeline step 13 | `build-blog-data.yml` -> Cloudflare R2 -> the blog |
| `ratings-data` | `career_panna.parquet`, `career_panna_asof.parquet` | Skills pipeline steps 09/09b | Predictions pipeline (panna feature) |

### The asset-updatedAt gotcha

**A release's `createdAt`/`publishedAt` reflects when the tag was first
created, not when its assets were last uploaded.** Workflows reuse the same
tag and re-upload files under it repeatedly, so a release can show as months
old while the data inside is from this morning. Always check **asset-level**
`updatedAt`, not the release page's date. This is exactly what the `vb_*`
introspection helpers do for you:

```{r asset-freshness}
assets <- vb_list_assets("peteowen1/pannadata", "opta-latest")
max(assets$updated_at)                                   # true freshness
vb_generation("peteowen1/pannadata", "opta-latest")       # same idea, one call
```

## Publishing: `vb_publish()`

If you change a pipeline step that produces a published artifact, you
publish it with `vb_publish()` -- a manifest-gated, hash-first upload that
never leaves a release half-updated. The real call site, simplified from
`data-raw/match-predictions-opta/13_publish_release_data.R`:

```{r publish}
blog_files <- c("data-raw/cache-predictions-opta/panna_ratings.parquet",
                "data-raw/cache-predictions-opta/match_predictions.parquet")

manifest <- vb_publish(blog_files, repo = "peteowen1/pannadata", tag = "blog-latest")
manifest$generation
```

Order of operations inside `vb_publish()`: hash the files, upload with
bounded retries collecting any failures, verify the live asset list, and
only THEN write `bus_manifest.json` -- last. If any upload fails, the
manifest write never happens, so consumers keep seeing the last consistent
snapshot instead of a torn mix of old-and-new files. Useful arguments:

- `carry_forward = TRUE` (default) -- merge with the previous manifest so a
  partial publish (e.g. only game logs changed) still describes the whole tag.
- `min_row_frac` -- optional floor vs. the previous manifest's row counts,
  to catch an accidentally-truncated export before it goes live.
- `dry_run = TRUE` -- build and return the manifest without uploading anything.

## Data conventions

- **Parquet or CSV only** -- never RDS for published data (RDS is fine for
  local pipeline caches, which stay out of git entirely).
- **Data and trained models never go into git** -- everything above moves
  through GitHub Releases, not commits.
- **Small published tables (<100KB / <10k rows)** ship as parquet *and* CSV;
  larger ones (game logs, action equity) ship parquet only.

## Loading published data

Once data is downloaded (or without downloading it at all -- every loader
defaults to `source = "remote"`), the `load_opta_*()` family is the read
side:

```{r load-family}
stats    <- load_opta_stats("ENG", season = "2024-2025")
skills   <- load_opta_skills()
psr_wk   <- load_opta_psr_weekly(date = "2026-03-18")
preds    <- load_predictions(source = "remote")
```

Under the hood, remote reads go through an internal column-pruned parquet
reader so a call like `load_opta_stats("ENG", columns = c("player_name",
"goals"))` only pulls the columns you asked for over the network rather than
the full 263-column table -- you don't call that reader directly, just pass
`columns` to whichever `load_opta_*()` function you're using.

## Next steps

- [Pipeline Anatomy](pipeline-walkthrough.html) -- what produces each published artifact
- [Player Ratings](player-ratings.html) -- the metrics these artifacts contain
- [Getting Started](getting-started.html) -- installation and first data load
