---
title: "Getting Started with panna"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with panna}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This vignette covers installation, data download, and basic usage of the panna package.

## Installation

Install panna from GitHub:

```{r install, eval = FALSE}
# Install devtools if needed
install.packages("devtools")

# Install panna
devtools::install_github("peteowen1/panna")
```

## Downloading Data

Data is stored in GitHub Releases. Most `load_*()` functions default to `source = "remote"`, so you can start using data immediately. For faster repeated access, download data locally first:

```{r download, eval = FALSE}
library(panna)

# Download Opta data (the project's sole data source)
pb_download_source("opta")
```

Data is cached locally after download, so you only need to do this once (or when updating to get new matches).

## Loading Data

### Opta Data

Opta is the package's data source, with 263 columns per player across 15 leagues. It powers the full RAPM/SPM pipeline, EPV, xMetrics, estimated skills, and match predictions.

```{r load-opta, eval = FALSE}
# Load Opta player stats
opta_stats <- load_opta_stats("EPL", "2024-2025")

# Load Opta shots
opta_shots <- load_opta_shots("EPL", "2024-2025")

# Load all Big 5 leagues at once
big5 <- load_opta_big5("2024-2025")

# Additional Opta data types
match_events <- load_opta_match_events("EPL", "2024-2025")  # All events with x/y
lineups <- load_opta_lineups("EPL", "2024-2025")             # Lineup data
fixtures <- load_opta_fixtures("EPL")                        # Fixtures/results
xmetrics <- load_opta_xmetrics("EPL", "2024-2025")           # Pre-computed xG/xA/xPass
```

## League and Season Codes

### Opta League Codes (15 leagues)

| League | Opta Code | Season Format |
|--------|-----------|---------------|
| Premier League | EPL | 2024-2025 |
| La Liga | La_Liga | 2024-2025 |
| Bundesliga | Bundesliga | 2024-2025 |
| Serie A | Serie_A | 2024-2025 |
| Ligue 1 | Ligue_1 | 2024-2025 |
| Eredivisie | NED | 2024-2025 |
| Primeira Liga | POR | 2024-2025 |
| Super Lig | TUR | 2024-2025 |
| Championship | ENG2 | 2024-2025 |
| Scottish Premiership | SCO | 2024-2025 |
| Champions League | UCL | 2024-2025 |
| Europa League | UEL | 2024-2025 |
| Conference League | UECL | 2024-2025 |
| World Cup | WC | 2018 Russia |
| Euros | EURO | 2024 Germany |

### Season Formats

| Type | Format | Example |
|--------|--------|---------|
| Domestic leagues | YYYY-YYYY | "2024-2025" |
| Tournaments | YYYY Country | "2018 Russia" |

## Aggregated Player Statistics

The package provides convenience functions for aggregated player statistics:

```{r player-stats, eval = FALSE}
# Opta aggregated summary stats
opta_players <- player_opta_summary(
  leagues = "EPL",
  seasons = "2024-2025",
  min_minutes = 900  # Filter for players with 900+ minutes
)

# Other aggregated stat tables (same arguments)
players_passing <- player_opta_passing(
  leagues = c("EPL", "La_Liga"),               # Multiple leagues
  seasons = c("2023-2024", "2024-2025"),       # Multiple seasons
  min_minutes = 1800
)

players_defense <- player_opta_defense(leagues = "EPL", seasons = "2024-2025")
players_shots   <- player_opta_shots(leagues = "EPL", seasons = "2024-2025")
```

## Working with Multiple Leagues

Many functions accept vectors of leagues and seasons:

```{r multi-league, eval = FALSE}
# Load Big 5 Opta stats
big5_opta <- load_opta_big5("2024-2025")

# Or load specific leagues individually
multi_league <- data.table::rbindlist(lapply(
  c("EPL", "La_Liga", "Bundesliga"),
  function(lg) load_opta_stats(lg, "2024-2025")
), fill = TRUE)
```

## Advanced Features

Beyond basic data loading and player statistics, panna includes several advanced analytical pipelines:

- **EPV (Expected Possession Value)**: Action-level player valuation from Opta event data with x/y coordinates. See `?calculate_action_epv`.
- **Estimated Skills**: Bayesian decay-weighted skill estimation with position-specific priors and context adjustments. See `?estimate_player_skills` and `?player_skill_profile`.
- **Match Predictions**: XGBoost Poisson/multinomial models for match outcome prediction using Elo, rolling form, and team-level skill features. See `?predict_match`.

These pipelines are built on top of Opta data and require running the RAPM/SPM pipeline first.

## Next Steps

- [Player Ratings](player-ratings.html) - Learn about RAPM and SPM methodology
- [Data Sources](data-sources.html) - Opta league codes, season formats, and loaders
- [Data Dictionary](../DATA_DICTIONARY.md) - Column definitions
