---
title: "How the panna EPV model works"
subtitle: "From raw football events to 'who actually added value?'"
author: "Pete Owen"
date: today
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: true
    code-summary: "Show the code"
    theme: cosmo
    fig-width: 9
    fig-height: 5.5
    embed-resources: true
execute:
  warning: false
  message: false
---

::: {.callout-note}
## The one-paragraph version

Every moment a team has the ball is worth something — a throw-in deep in your
own half is worth almost nothing, the ball at a striker's feet in the box is
worth a lot. **EPV (Expected Possession Value)** puts a number on that worth at
every instant, using a machine-learning model trained on millions of actions.
Every pass, dribble, duel and shot *changes* that number, and the change is
credited to the players involved. Sum the credits over a season and you get a
measure of who actually adds value — and *how* they add it.
:::

```{mermaid}
%%| fig-cap: "The pipeline at a glance."
flowchart LR
  A["Opta event feeds<br/>(daily scrape)"] --> B["SPADL actions<br/>(standardised language)"]
  B --> C["Possession chains<br/>+ outcome labels"]
  C --> D["XGBoost EPV model<br/>(value every action)"]
  D --> E["Credit assignment<br/>(who caused the change?)"]
  E --> F["Player value profiles<br/>(season aggregates)"]
```

```{r}
#| label: setup
#| output: false
devtools::load_all("../..", quiet = TRUE)   # the panna package
suppressPackageStartupMessages({
  library(data.table)
  library(ggplot2)
})
theme_set(theme_minimal(base_size = 12))

# One real match, frozen as a test fixture: West Ham 3-0 Leeds, 2026-05-24
fixtures <- "../../tests/testthat/fixtures"
events  <- as.data.table(arrow::read_parquet(file.path(fixtures, "opta_match_events_epl.parquet")))
labeled <- as.data.table(arrow::read_parquet(file.path(fixtures, "spadl_labeled_epl.parquet")))
```

# Step 1 — The raw data

Everything starts with **Opta event feeds**: a row for every touch in a match,
with a type code, pitch coordinates, an outcome flag and a timestamp. A scraper
runs every morning on GitHub Actions, pulls the latest matches for ~100
competitions and publishes them as parquet files. One Premier League season is
roughly **650,000 events across 380 matches**.

This whole document uses a single real match — **West Ham 3–0 Leeds
(24 May 2026)** — which arrives as `r nrow(events)` raw event rows.
Here is what the machine actually sees (these rows are West Ham's first goal):

```{r}
#| label: tbl-raw
#| tbl-cap: "Raw Opta events around a goal. type_id 16 = goal, 1 = pass."
goal_idx <- which(events$type_id == 16)[1]
knitr::kable(
  events[(goal_idx - 3):goal_idx,
         .(type_id, minute, second, player_name, x, y, outcome)],
  digits = 1
)
```

Cryptic type codes, coordinates as percentages, no sense of what an action *led
to*. Not something you can model directly — first it needs a common language.

One quirk worth knowing before any pitch plot: coordinates run 0–100 as
percentages of the pitch, and they are **team-relative** — every team's actions
are recorded as if attacking toward x = 100. You can see it in the table above:
the defender's clearance sits at x ≈ 5 while the attacker's pass next to it is
at x ≈ 95 — the *same patch of grass*, described from two perspectives. It also
means every pitch plot in this document reads left → right no matter who has
the ball.

# Step 2 — A standard language for actions (SPADL)

We convert raw events to **SPADL** (Soccer Player Action Description Language):
one row per *deliberate on-ball action*, with a readable action type, start and
end coordinates, the result, and who was involved — including the opponent in
duels. Referee bookkeeping (cards, subs, stoppages) is filtered out, and the
two halves of a contested duel are merged into a single winner-vs-loser row.

```{r}
#| label: tbl-spadl
#| tbl-cap: "The same goal, now as SPADL actions. The last column is the target the model will learn (more in Step 3): the xG of the next shot in the half, signed + if your own team takes it, − if the opponent does."
spadl_cols <- c("action_id", "time_seconds", "player_name", "action_type",
                "start_x", "start_y", "end_x", "end_y", "result", "bodypart",
                "next_xg_label")
goal_action <- labeled[action_type == "shot" & result == "success"][1]
knitr::kable(
  labeled[action_id %between% c(goal_action$action_id - 3, goal_action$action_id),
          ..spadl_cols],
  digits = c(0, 0, 0, 0, 1, 1, 1, 1, 0, 0, 3)
)
```

Notice the sign of `next_xg_label` flipping between the two teams' actions in
the build-up: the same upcoming shot is *good news* for one side and *bad news*
for the other.

The `r nrow(events)` raw events become `r nrow(labeled)` SPADL actions.
Actions are then grouped into **possession chains** — an unbroken spell of one
team having the ball, which ends when the ball is turned over, goes out of
play, or someone shoots. This match had `r uniqueN(labeled$chain_id)` chains, of
which `r uniqueN(labeled[chain_outcome == "goal"]$chain_id)` ended in a goal
(it finished 3–0, so that's a good sign).

```{r}
#| label: fig-chain
#| fig-cap: "Each numbered label is one action in the chain: who did it and what it was. Arrows show where the ball travelled; everything attacks left to right."
goal_chains <- labeled[chain_outcome == "goal", .N, by = chain_id][order(-N)]
best_chain <- goal_chains$chain_id[1]

chain_steps <- labeled[chain_id == best_chain][order(action_id)]
chain_steps[, step := .I]
chain_steps[, minute := floor(time_seconds / 60) + fifelse(period_id == 2L, 45L, 0L)]
scorer <- chain_steps[.N, player_name]

plot_chain(labeled, target_match_id = labeled$match_id[1],
           target_chain_id = best_chain) +
  ggrepel::geom_text_repel(
    data = chain_steps,
    aes(start_x, start_y,
        label = sprintf("%d. %s\n%s", step, player_name, action_type)),
    size = 3.2, fontface = "bold", lineheight = 0.95,
    box.padding = 0.7, point.padding = 0.4,
    min.segment.length = 0, seed = 42
  ) +
  labs(
    title = sprintf("West Ham's goal, finished by %s (%d′)",
                    scorer, chain_steps[.N, minute]),
    subtitle = sprintf(
      "One possession chain: %d actions from winning the ball to scoring",
      nrow(chain_steps))
  )
```

# Step 3 — Putting a value on every action

Here is the core idea. For every action we ask the model:

> *Given where the ball is, what just happened, and how much time is left —
> how much shot danger is coming next, and for which team?*

Formally, the **target** the model learns is the expected-goals (xG) value of
the **next shot in the half**, signed: positive if your team takes it, negative
if the opponent does, zero if nobody shoots again. Because xG is a probability,
the target lives between **−1 and +1** — though in practice the extremes are
unreachable (a penalty is ~0.79, the very best chances ~0.9) and most actions
sit near zero. Train that on millions of actions and the prediction becomes
the EPV: the value of possession *right now*.

The unit, throughout everything that follows, is **goals**. A typical action
carries an EPV of about 0.02 — two hundredths of a goal of upcoming shot
danger — while possession in the heart of the box is worth 0.2–0.3. The
*changes* are smaller still: even an excellent line-breaking pass adds a few
hundredths of a goal. That's the right intuition for this whole framework —
single actions are tiny, and players separate themselves by stacking thousands
of them.

```{r}
#| label: epv-calc
#| output: false
epv_model   <- load_epv_model()
xg_model    <- load_xg_model()
xpass_model <- load_xpass_model()

valued   <- calculate_action_epv(labeled, epv_model = epv_model,
                                 xg_model = xg_model, league = "EPL")
credited <- as.data.table(assign_epv_credit(valued, xpass_model = xpass_model))
```

```{r}
#| label: fig-epv-pitch
#| fig-cap: "Every action in the match, coloured by its EPV. Value is low everywhere except the final third — and explodes near the box."
pitch_dt <- credited[!is.na(epv)]
draw_pitch() +
  geom_point(data = pitch_dt,
             aes(x = start_x, y = start_y, colour = pmin(epv, 0.3)),
             size = 1.6, alpha = 0.7) +
  scale_colour_viridis_c(option = "inferno", name = "EPV") +
  labs(title = "Where is possession worth something?",
       subtitle = "All actions attack left → right")
```

What the model cares about is not the value itself but the **change** each
action causes — `epv_delta`. A pass that breaks a defensive line jumps the EPV
up; a sloppy touch that leads to a turnover sends it negative. Watch the value
climb through the goal chain from @fig-chain:

```{r}
#| label: fig-epv-timeline
#| fig-cap: "EPV through the goal chain: each action moves the number until the shot cashes it in."
chain_dt <- credited[chain_id == best_chain][order(action_id)]
chain_dt[, step := .I]
ggplot(chain_dt, aes(step, epv)) +
  geom_step(linewidth = 0.9, colour = "grey40") +
  geom_point(aes(colour = action_type), size = 3) +
  geom_text(aes(label = player_name), angle = 35, hjust = -0.05,
            vjust = -0.6, size = 3) +
  scale_x_continuous(breaks = chain_dt$step, labels = chain_dt$action_type,
                     expand = expansion(mult = c(0.05, 0.18))) +
  scale_y_continuous(expand = expansion(mult = c(0.1, 0.25))) +
  labs(x = NULL, y = "EPV (expected next-shot xG, signed)") +
  theme(axis.text.x = element_text(angle = 35, hjust = 1),
        legend.position = "none")
```

# The model itself

There is deliberately no deep learning here — the model is an **XGBoost
gradient-boosted tree regressor**, trained on ~5 million actions sampled across
every league we rate. It sees 15 features per action:

```{r}
#| label: tbl-features
#| tbl-cap: "The 15 features the EPV model sees for every action."
features <- data.frame(
  Feature = c("start_x, start_y", "distance_to_goal", "angle_to_goal",
              "dx, dy", "time_remaining", "prev_x, prev_y",
              "prev_dx, prev_dy", "same_team_prev", "action_cat",
              "result_success", "league_id"),
  `Plain English` = c(
    "Where on the pitch the action starts",
    "How far from the goal",
    "How wide the goal looks from here",
    "How far the ball travels (direction matters)",
    "Seconds left in the half",
    "Where the previous action happened",
    "How the previous action moved the ball",
    "Did my team also have the previous touch?",
    "What kind of action this is (pass / dribble / shot / ...)",
    "Did the action succeed?",
    "Which league we are in (a Premier League box touch ≠ a League Two one)"
  ),
  check.names = FALSE
)
knitr::kable(features)
```

And what it actually leans on, from the trained model:

```{r}
#| label: fig-importance
#| fig-cap: "Feature importance from the production model: location dominates, exactly as football intuition says it should."
imp <- xgboost::xgb.importance(model = epv_model$model)
ggplot(as.data.table(imp)[order(Gain)][, Feature := factor(Feature, levels = Feature)],
       aes(Gain, Feature)) +
  geom_col(fill = "#2c7fb8") +
  labs(x = "Share of model gain", y = NULL)
```

::: {.callout-tip}
## Why "next shot xG" and not "next goal"?

Goals are rare (about 3 per match) and noisy — a model trained on them mostly
learns luck. Shots happen ~25 times a match and their xG is a well-calibrated
danger measure, so the signal per action is ~10x richer. This is the same trick
the analytics community uses everywhere: model the *process*, not the outcome.
:::

::: {.callout-important}
## Isn't `result_success` cheating? (the leakage question)

Seeing the *outcome* of an action in the feature list looks suspicious — isn't
that information from the future? It isn't, because of the order things happen
in. Leakage means using information that arrives **after the thing you're
predicting**. Here the timeline is:

> action happens → **its result resolves** → ... play continues ... →
> **the next shot** (the target)

The result of the current action is settled *before* the target exists, so the
model is allowed to know it. What it's really doing is defining the game state
properly: a pass *completed* into the box and the same pass *intercepted* are
two completely different situations, and the EPV must differ — that difference
is exactly what generates the negative `epv_delta` that blames the player who
lost the ball. Strip the result out and a turnover would look identical to a
line-breaking pass, and credit assignment would fall apart. (This is the same
design as the academic VAEP framework, which values the state *after* each
action.)

What **would** be leakage: features like "did this possession chain end in a
goal" or anything derived from the next shot itself — those live on the label
side of the data and never enter the feature matrix. One related subtlety:
shots don't even use the model's prediction — a shot's EPV is set directly to
its xG, so the model never gets to peek at whether a shot became a goal.
:::

# Step 4 — Giving the value to players

An EPV change has to be *earned by someone*. There are four credit rules, and
each one is below, caught in the act in this match — the row marked **→** is
the action the rule applies to, with the action before and after for context.

```{r}
#| label: credit-helper
# Show an action with its neighbours, numeric columns rounded
show_window <- function(idx, cols) {
  rows <- seq(max(idx - 1L, 1L), min(idx + 1L, nrow(credited)))
  win <- credited[rows, ..cols]
  num <- names(win)[vapply(win, is.numeric, logical(1))]
  win[, (num) := lapply(.SD, round, 3), .SDcols = num]
  win[, ` ` := data.table::fifelse(rows == idx, "→", "")]
  data.table::setcolorder(win, " ")
  win
}
```

**Normal actions** — the acting player simply keeps the full `epv_delta` their
action caused. Here, the match's most valuable completed pass:

```{r}
#| label: tbl-credit-normal
#| tbl-cap: "player_credit equals epv_delta — nothing to share."
cand <- credited[, which(action_type == "pass" & result == "success" &
                           possession_change == FALSE & !is.na(epv_delta))]
i_normal <- cand[which.max(credited$epv_delta[cand])]
knitr::kable(show_window(i_normal,
  c("player_name", "action_type", "result", "epv", "epv_delta", "player_credit")))
```

**Turnovers** — when possession flips, the `epv_delta` is negative for the team
that had the ball. Blame is split: half to the player who lost it
(`player_credit`), and the mirror image to the opponent who won it
(`receiver_credit`). Why exactly half? Because event data can't tell us whether
the moment was the passer's mistake or the defender's brilliance — with no way
to know, an even split is the honest default. The match's costliest giveaway:

```{r}
#| label: tbl-credit-turnover
#| tbl-cap: "Blame shared between the player losing the ball and the opponent gaining it."
cand <- credited[, which(possession_change == TRUE & epv_delta < 0 &
                           receiver_credit > 0)]
i_turn <- cand[which.min(credited$epv_delta[cand])]
knitr::kable(show_window(i_turn,
  c("player_name", "action_type", "result", "epv_delta",
    "player_credit", "receiver_player_name", "receiver_credit")))
```

**Duels** — aerials and take-on-vs-tackle contests are merged into one row, and
credit is zero-sum: the winner (`player_credit`) gains exactly what the loser
(`opponent_credit`) loses. The match's biggest aerial win:

```{r}
#| label: tbl-credit-duel
#| tbl-cap: "Zero-sum: winner's gain + loser's loss = 0."
cand <- credited[, which(action_type == "aerial" & !is.na(opponent_credit))]
i_duel <- cand[which.max(abs(credited$player_credit[cand]))]
knitr::kable(show_window(i_duel,
  c("player_name", "action_type", "result", "player_credit",
    "opponent_player_name", "opponent_credit")))
```

**Saved shots** — the shooter keeps the value of having created the attempt;
the keeper earns separate credit for taking it away (`receiver_credit`). The
biggest save of the match:

```{r}
#| label: tbl-credit-save
#| tbl-cap: "The shooter is not punished for a good attempt; the keeper is rewarded for stopping it."
cand <- credited[, which(opta_type_id == 15 & receiver_credit > 0)]
i_save <- cand[which.max(credited$receiver_credit[cand])]
knitr::kable(show_window(i_save,
  c("player_name", "action_type", "result", "epv_delta",
    "player_credit", "receiver_player_name", "receiver_credit")))
```

The goal chain again, now with credit attached:

```{r}
#| label: tbl-credit
#| tbl-cap: "Per-action EPV credit for the goal chain. The scorer cashes in the biggest jump, but the build-up passes earn real value too."
knitr::kable(
  chain_dt[, .(player_name, action_type, result,
               epv = round(epv, 3), epv_delta = round(epv_delta, 3),
               player_credit = round(player_credit, 3))]
)
```

# Step 5 — A season of credit: player value profiles

Run that over every action of a season, sum by player and action type,
normalise per 90 minutes — and each player gets a *profile* of where their
value comes from.

Two of the areas below never appeared in the SPADL action list, because they
come from the credit *channels* rather than the action itself:
**possession_won** is the receiver-side credit from the turnover rule — being
the player who claims the ball when the opponent gives it away — and
**shot_stop** is the save/block credit from the saved-shot rule. The units are
still goals: a shooting value of 0.12 per 90 means roughly 0.12 goals of shot
danger added per match through shooting.

```{r}
#| label: load-season
season_path <- "../../debug/player_value_by_area_EPL_2025-2026.parquet"
has_season <- file.exists(season_path)
```

```{r}
#| label: fig-profiles
#| eval: !expr has_season
#| fig-cap: "Two very different ways to be valuable. Haaland's value is concentrated in shooting; Van Dijk's comes from what he does with the ball, not his (famously good) aerial duels — defensive-half aerials just don't swing EPV much."
vba <- as.data.table(arrow::read_parquet(season_path))
two <- vba[player_name %in% c("E. Haaland", "V. van Dijk")]

# union of each player's top-7 areas, shown in ONE shared order (and shared
# x scale) so the two profiles can be compared row by row
top_areas <- unique(two[order(-credit_p90), head(.SD, 7), by = player_name]$area)
two <- two[area %in% top_areas]
area_order <- two[, .(total = sum(credit_p90)), by = area][order(total), area]
two[, area := factor(area, levels = area_order)]

ggplot(two, aes(credit_p90, area, fill = player_name)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~player_name) +
  scale_fill_manual(values = c("#6CABDD", "#C8102E")) +
  labs(x = "EPV credit per 90 minutes", y = NULL,
       title = "Where does each player's value come from? (EPL 2025-26)",
       subtitle = "Same areas, same order, same scale — read across to compare")
```

```{r}
#| label: tbl-leaders
#| eval: !expr has_season
#| tbl-cap: "Area leaders, EPL 2025-26 (900+ minutes): shooting value vs aerial value. Different skills, different players — which is the whole point."
qual <- vba[minutes >= 900]
leaders <- rbind(
  qual[area == "shot"][order(-credit_p90)][1:5][, skill := "Shooting"],
  qual[area == "aerial"][order(-credit_p90)][1:5][, skill := "Aerial duels"]
)
knitr::kable(
  leaders[, .(skill, player_name, minutes, n_actions,
              credit_p90 = round(credit_p90, 3))]
)
```

# The questions you should be asking

Two of the best ones — *why next-shot xG instead of goals* and *isn't
`result_success` leakage* — are answered up in the model section. Here are the
others, click to expand.

::: {.callout-note collapse="true"}
## How do you know the model is any good?

Three layers. First, standard ML hygiene: it's trained with 5-fold
cross-validation and early stopping, so it's always judged on actions it never
saw. Second, sanity: its predictions reproduce things we know are true (the
feature-importance chart and the pitch heatmap both recover football's
geography of danger without being told it). Third — and really the one that
matters — the downstream test: ratings built from these numbers feed a match
prediction model whose forecasts are scored against real results every week.
A value model that didn't measure real value would wash out there. Any single
action's EPV is noisy; the reliability comes from averaging thousands of them.
:::

::: {.callout-note collapse="true"}
## Does the model know the score?

No — look back at the feature list: clock yes, score no. That's deliberate. A
team 3–0 up stops attacking, and if the model conditioned on score, players in
blowouts would have their actions systematically re-valued by context they
don't control. Keeping EPV score-blind makes it a measure of the *action*;
score and clock pressure are handled by a separate win-probability model (WPA),
and the two views are blended later. The honest cost: EPV can't tell a
meaningful 88th-minute pass from a garbage-time one — that's exactly the gap
WPA exists to fill.
:::

::: {.callout-note collapse="true"}
## Couldn't a striker farm value by shooting from anywhere?

No — shooting is valued against the situation the shot came *from*. A shot's
EPV is its xG, so blasting one from 30 yards (xG ≈ 0.03) out of a possession
worth 0.10 *destroys* value, and a miss hands the full negative delta to the
shooter as blame. The shooting leaderboard rewards shot **selection and
finishing position**, not volume. If anything the punishment is the thing to
watch: speculative shooters sit *below* quieter players who only shoot from
good spots.
:::

::: {.callout-note collapse="true"}
## Why XGBoost and not a neural network?

Because the data is 15 tabular features by a few million rows, and on tabular
data gradient-boosted trees are still the thing to beat. The practical wins
matter too: it retrains in minutes on a free CI runner every week, needs no
GPU, and gives interpretable feature importances. Deep learning earns its
complexity with *tracking* data — 22 players' coordinates 25 times a second —
which is a different (and much more expensive) data world than event feeds.
:::

::: {.callout-note collapse="true"}
## What can the model not see at all?

Everything off the ball. Event data records ~1,250 touches per match and
nothing else: no runs that drag defenders away, no marking, no positioning
that makes the dangerous pass never get attempted. Two mitigations are baked
in. Defensive on-ball actions get a deliberate 1.5× credit boost, because
defence is structurally under-represented in touch data — a great tackle often
appears in the data as *the attacker's* turnover. And the headline panna
rating doesn't rely on EPV alone: it blends in RAPM, a plus-minus model that
infers off-ball impact from how the score moves while a player is on the
pitch. EPV tells you what a player did with the ball; RAPM catches what
happened *because they were there*.
:::

# Where this goes next

Per-action EPV credit is the *foundation*, not the finished rating. Downstream,
the same numbers feed:

- **EPR** — a regression that turns game-level EPV into a stable player
  rating, controlling for league strength and opponent quality;
- **WPA** — the same credit idea applied to *win probability* instead of shot
  danger, so late-game moments weigh what they should;
- **panna** — the headline rating, which blends these with a plus-minus model
  (RAPM) that catches everything off-ball that event data can't see.

All of it published weekly, automatically, from the same pipeline you just
walked through.

::: {.callout-note collapse="true"}
## Reproducing this document

Everything here runs from the panna repo: the single-match data is a committed
test fixture, the models download themselves on first use, and the season
table comes from `debug/player_value_by_area.R`. Render with
`quarto render data-raw/explainer/epv_explainer.qmd` from the `panna/` root.
:::
